### 2022-06-07 src/ms.tex
linear regression section

% Math 
\usepackage{amsmath,amssymb,bm}
\DeclareMathOperator*{\argmax}{arg\,max}
\usepackage{siunitx} % for proper units 
\usepackage{esint} % various fancy integral symbols
\usepackage[bbgreekl]{mathbbol}

@@

In general, we are usually dealing with several parameters at once in which
case we are interested in \emph{joint} probability distributions over all the
parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
density function $p(x_1,x_2)$ which when integrated over a particular interval
gives the probability that both $x_1$ and $x_2$ are contained within that
interval. These joint probabilities can be decomposed into \emph{conditional
    probabilities} using the so called product rule:
\begin{align}
    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
    \label{eq:product_rule}
\end{align}
where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
\begin{equation}
    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
    \label{eq:bayes_theorem}
\end{equation}
Bayes's theorem is simply a rule for converting one kind of a conditional probability
into another.
Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
\emph{marginalize} parameters that we are not interested in to obtain a distribution over
parameters of interest\footnote{
    It is very common in astronomy to have a model
    which consists of a set of parameters $\boldsymbol x_1$ of physical signal of interest and another set
    of ``nuisance parameters'' $\boldsymbol x_2$ which often describe the noise properties. In those cases
    we need to marginalize over parameters $\boldsymbol x_2$ to obtain a distribution over $x_1$.}:
\begin{equation}
    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
    x_2)\,p(x_2)\textrm{d}x_2
\end{equation}
where we applied Bayes's theorem in the second inequality.
In cases where the parameter space is high dimensional these integrals might be
impossible to calculate analytically but they can (sometimes) be estimated using sampling
algorithms as we shall se in subsequent sections.
In some cases, we may be able to factor out $p(x_1,x_2)$ as
If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
\begin{equation}
    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
\end{equation}
Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
\begin{equation}
    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
    \label{eq:likelihood_indep}
\end{equation}
If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
identically distributed (``iid''). If the data is independent but not identically distributed
we say that the uncertainties are \emph{heteroscedastic}.
The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
uncertainties.

@@

\subsection{Maximum likelihood estimation}
One of the most popular inference methods is called \emph{maximum likelihood
    estimation} (MLE). Maximum likelihood estimation answers the question:
\textbf{what choice of parameters $\vec{x}$ makes my data most likely} as
opposed to the question \textbf{what are the most likely values of the
    parameters $\vec{x}$ having observed the data $\mathcal{D}$}\footnote{The
    distinction between the two questions is subtle, but important. The parameters
    which maximize the likelihood might be very different from those which are most
    ``sensible'' and which best describe the data.} which necessitates the
multiplication of the likelihood with a prior. Since we don't incorporate the
prior, it is not a Bayesian method. The output of MLE is a \emph{point
    estimate} of the parameters $\hat{\vec{\theta}}$. In a Bayesian viewpoint, the
MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.

\subsubsection{MLE for a Gaussian linear model}

\subsection{Gaussian linear models}

>>>>

% Math 
\usepackage{amsmath,amssymb,bm}
\DeclareMathOperator*{\argmax}{arg\,max}
\usepackage{siunitx} % for proper units 
\usepackage{esint} % various fancy integral symbols
%\usepackage[bbgreekl]{mathbbol}

@@

In general, we are usually dealing with several parameters at once in which
case we are interested in \emph{joint} probability distributions over all the
parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
density function $p(x_1,x_2)$ which when integrated over a particular interval
gives the probability that both $x_1$ and $x_2$ are contained within that
interval. These joint probabilities can be decomposed into \emph{conditional
    probabilities} using the so called product rule:
\begin{align}
    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
    \label{eq:product_rule}
\end{align}
where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
\begin{equation}
    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
    \label{eq:bayes_theorem}
\end{equation}
Bayes's theorem is simply a rule for converting one kind of a conditional probability
into another.
Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
\emph{marginalize} parameters that we are not interested in to obtain a distribution over
parameters of interest:
\begin{equation}
    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
    x_2)\,p(x_2)\textrm{d}x_2
\end{equation}
where we applied Bayes's theorem in the second inequality.
In cases where the parameter space is high dimensional these integrals might be
impossible to calculate analytically but they can (sometimes) be estimated using sampling
algorithms as we shall se in subsequent sections.
In some cases, we may be able to factor out $p(x_1,x_2)$ as
If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
\begin{equation}
    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
\end{equation}
Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
\begin{equation}
    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
    \label{eq:likelihood_indep}
\end{equation}
If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
identically distributed (``iid''). If the data is independent but not identically distributed
we say that the uncertainties are \emph{heteroscedastic}.
The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
uncertainties.

@@

\subsection{Maximum likelihood estimation}
One of the most popular inference methods is called \emph{maximum likelihood
    estimation} (MLE). Maximum likelihood estimation answers the question:
\textbf{what choice of parameters $\vec{x}$ makes my data most likely} as
opposed to the question \textbf{what are the most likely values of the
    parameters $\vec{x}$ having observed the data $\mathcal{D}$}\footnote{The
    distinction between the two questions is subtle, but important. The parameters
    which maximize the likelihood might be very different from those which are most
    ``sensible'' and which best describe the data.} which necessitates the
multiplication of the likelihood with a prior. Since we don't incorporate the
prior, it is not a Bayesian method. The output of MLE is a \emph{point
    estimate} of the parameters $\hat{\vec{\theta}}$. In a Bayesian viewpoint, the
MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.
The MLE estimate is obtained by maximizing the logarithm\footnote{Because
    probabilities in most region of the parameter space are very small numbers we
    always use the log transform when optimizing likelihoods or posteriors.} of the
likelihood $p(\mathcal{D}\lvert\vec{\theta})$ (or equivalently, minimizing the
negative log likelihood) with respect to the parameters $\vec{\theta}$:
\begin{equation}
    \hat{\boldsymbol{\theta}}_{\mathrm{mle}}=\underset{\boldsymbol{\theta}}{\operatorname{argmax}}
    \ln p\left(\mathcal{D}\lvert \boldsymbol{\theta}\right)
\end{equation}

\subsubsection{MLE for a multivariate Gaussian}
The log-likelihood of multivariate Gaussian distribution without the
unnecessary constants is
\begin{equation}
    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\log p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \log |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\top} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
\end{equation}
where $\vec{y}_n$ are the data points and $\vec{\Sigma}^{-1}$ is
the inverse covariance matrix, also knowns as the \emph{precision matrix}.
The MLE for the mean parameter $\hat{\vec{\mu}}$
is given by the sample mean (empirical mean) of the data \citep{murphy_book_2022}:
\begin{equation}
    \hat{\boldsymbol{\mu}}=\frac{1}{N} \sum_{n=1}^{N} \boldsymbol{y}_{n}=\overline{\boldsymbol{y}}\quad .
\end{equation}
Similarly, the MLE of the the covariance matrix $\hat{\vec{\Sigma}}$ is equal to
the empirical covariance matrix
\begin{equation}
    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\top}
\end{equation}
Thus, whenever we compute an empirical mean and the covariance for a given datasets we
are implicitly assuming that the data is normally distributed and obtaining an MLE
estimate of that distribution.

\subsection{Least squares linear regression}
In this section we cover the classical non Bayesian formulation of linear
regression, also known as \emph{least squares regression} and in the next
section we generalize to the Bayesian case. In a (multiple) linear regression
model we have (noisy) measurements $\vec{y}\in \mathbb{R}^n$ (\emph{dependent
    variable}), an $n\times p$ matrix $\vec{M}\in \mathbb{R}^{n\times p}$ of inputs
(\emph{independent variables}, \emph{covariates}, or \emph{features}) and a
vector of \emph{regression coefficients} or \emph{weights} $\vec{\theta}\in
    \mathbb{R}^p$. This is a very general model because the matrix $\vec{M}$ can
consist of basis functions such as polynomials, spherical harmonics, Fourier
modes, wavelets etc. so although the model is linear in the parameters
(regression coefficients), it is not generally linear in the raw input data. If
we assume that the observations $\vec{y}$ have Gaussian noise we can write down
the likelihood of this model as
\begin{equation}
    p(\vec{y}\lvert \vec{\theta})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta}, \vec{C})
\end{equation}
where $\vec{C}$ is the data covariance matrix and $\mathcal{N}$ is shorthand notation for
the multivariate normal distribution. In astronomical application it is usually
assumed to be diagonal and we might have some idea for the scale of the variance
of the diagonal elements (the ``errorbars'').
The negative log likelihood (NLL) is then equal (ignoring constants)  to
\begin{equation}
    \operatorname{NLL}(\vec{\theta})=\frac{1}{2}(\vec{M} \vec{\theta}-\vec{y})^{\top}\vec{C}^{-1}(\vec{M} \vec{\theta}-\vec{y})
\end{equation}
which is just the well known least squares objective or (half the) $\chi^2$ as it is known
in astronomy.

\subsubsection{Underdetermined case ($p < n$)}
For now, let's assume that the inverse data covariance matrix $\vec{C}^{-1}$ is
a diagonal matrix with the the same variance $\sigma^2$ for every data point.
In that case the MLE solution for the weights $\vec{\theta}$ is the solution to
the following (convex) optimization problem:
\begin{equation}
    \hat{\vec{\theta}}=
    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2}
    \label{eq:ols_objective}
\end{equation}
There is a closed form solution for Equation~\ref{eq:ols_objective} as long as the number
of parameters and features $p$ is less than the number of data points $n$ and the solution
is \citep{arXiv:2101.07256}
\begin{equation}
    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M})^{-1} \vec{M}^{\top} \vec{y}
    \label{eq:ols_solution}
\end{equation}
under the assumption that $\vec{M}\vec{M}^\top$ is invertible which is usually the case.
Equation~\ref{eq:ols_solution} is called the \emph{ordinary least squares} (OLS) solution.
For a nontrivial data covariance matrix $\vec{C}$ Equation~\ref{eq:ols_solution} generalizes
to
\begin{equation}
    \hat{\vec{\theta}}=(\vec{M}^{\top}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\top}\vec{C}^{-1} \vec{y}
    \label{eq:wls_solution}
\end{equation}
which is the solution to the \emph{weighted least squares} (WLS) problem.
It is common to modify the least squares objective functions with the addition of a
\emph{regularization} term to prevent overfitting. For instance, we can modify the OLS
objective (Equation~\ref{eq:ols_objective}) by adding an \emph{L2 regularization} term:
\begin{equation}
    \hat{\vec{\theta}}=
    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2
    \label{eq:ols_objective_l2}
\end{equation}
where $\lambda >0$ is a regularization parameter
which penalizes large magnitudes of the coefficients $\vec{\theta}$. This is also a convex
optimization problem and the solution is
\begin{equation}
    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\top} \vec{y}
    \label{eq:ols_solution_l2}
\end{equation}
where $\vec{I}$ is the identity matrix. Regularized OLS regression with this particular
regularization term is also known as \emph{ridge regression} in the statistics
literature\footnote{One of the many things that make statistics confusing is that very
    similar methods often have special names which obscures the big picture. The vast majority
    of statistical methods can be seen as a linear model with the data represented in a particular
    basis, and with a particular choice of the likelihood (noise model), priors (regularization)
    and  a method for obtaining the solution (usually a convex optimization problem).}
L2 regularization (as well as most other forms of regularization terms) can be seen
as a particular choice of priors for the regression coefficients in the Bayesian paradigm.
The above also naturally generalizes to the case of weighted least squares.

\subsubsection{Overdetermined case ($p > n$)}
Contrary to popular belief within the astronomy community it is possible and
often desirable to have a model with more parameters than data points (see
Chapter~\ref{ch:mapping_io} for a real world example). In the overparametrized
case of linear regression $(p > n)$ there are multiple choices for
$\vec{\theta}$ which perfectly fit the data. The solution in that case is
defined to be a minimum-norm parameter vector $\vec{\theta}$ that interpolates
the data ($\vec{y}=\vec{M} \vec{\theta}$), out of possibly many degenerate
solutions. The OLS objective in Equation~\ref{eq:ols_objective} can be modified
such that it is valid for both the $p<n$ and the $p>n$ case by making use of a
limit in which an L2 regularization term tends to zero
\citep{arXiv:2101.07256}:
\begin{equation}
    \hat{\vec{\theta}}=\lim _{\lambda \rightarrow 0^{+}}\left[
    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2\right]\quad .
\end{equation}
In this case the OLS solution becomes
\begin{equation}
    \hat{\vec{\theta}}=\vec{M}^{\top} (\vec{M} \vec{M}^{\top})^{-1} \vec{y}
    \label{eq:ols_solution_overdet}
\end{equation}
Equations~\ref{eq:ols_solution} and \ref{eq:ols_solution_overdet} can be unified into a
single equation by making use of the \emph{Moore-Penrose pseudoinverse} $\vec{M}^\dagger$.
The pseudoinverse is defined by taking the singular-value decomposition (SVD)
of $\vec{M}$:
\begin{equation}
    \vec{M}^\dagger\equiv\vec{U} \vec{S}^\dagger \vec{V}^\dagger
\end{equation}
Then $\vec{M}^\dagger\equiv  \vec{V}^{\top} \vec{S}^{\dagger} \vec{U}^{\top}$.

\subsection{Bayesian linear regression}
In the Bayesian paradigm we can write down the posterior over the regression
weights $\vec{\theta}$ assuming a Gaussian likelihood and a Gaussian prior over
the weights. The posterior is then proportional to
\begin{align}
    p(\vec{\theta}|\vec{y}) & \propto p(\vec{y}\lvert \vec{\theta})\,p(\vec{\theta})                                                          \\
                            & =\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta},\vec{C})\,\mathcal{N}(\vec{\theta}\lvert \vec{\mu},\vec{\Lambda})
    \label{eq:bayes_lin_regression}
\end{align}
where $\vec{C}$ is the data covariance matrix as before, $\vec{\mu}$ is the mean
of the Gaussian prior on coefficients and $\vec{\Lambda}$ is the the prior covariance matrix.
The product of two multivariate Gaussian distributions is also a multivariate Gaussian distribution
and the solution to Equation~\ref{eq:bayes_lin_regression} is \citep{arXiv:2005.14199}
\begin{equation}
    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{\theta}, \vec{C}) \,\mathcal{N}(\vec{\theta} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{\theta} \lvert\vec{a}, \vec{A}) \,\mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
\end{equation}
where
\begin{align}
     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\top} \vec{C}^{-1} \vec{M}                                                         \\
     & \vec{a}         =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\top} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
     & \mathbf{B}      =\mathbf{C}+\vec{M} \boldsymbol{\Lambda} \vec{M}^{\top}                                                        \\
     & \boldsymbol{b}  =\vec{M} \boldsymbol{\mu}
\end{align}
